# Analyses of randomized management response trials and observational surveys {#mrt}

*Edited by: Jordan Chamberlin, Markus Walsh, ...?*

## Introduction {#mrt-intro}

Inherently low soil fertility, nutrient imbalances and potentially accelerating soil physical, chemical and biological degradation, are thought to constitute threats to maintaining agricultural productivity and other ecosystem services across Africa's cropland biome. However, at present the geographical extent of existing soil problems in croplands, their location specific trends and opportunities for managing these over time are not known. An ability to track these potential changes and anticipate the attendant consequences requires reliable information about the characteristics and distribution of cropland productivity and the potentials for improvement, given any relevant interventions or pratices. 

The most reliable and informative way of assessing the efficacy and impact of management interventions is the use of [**Randomized Controlled Trials**](https://en.wikipedia.org/wiki/Randomized_controlled_trial) (RCT). However, the main problems with all RCTs is that they are usually logistically difficult and quite expensive to conduct.

... agricultural development projects frequently carry out regular monitoring and evaluation surveys to gauge impacts or to adjust management interventions where needed. More generally, these are known as [**Observational Surveys**](https://en.wikipedia.org/wiki/Observational_study) that can be applied in various configurations. Although observational surveys cannot be used to make definitive statements of fact about the "effectiveness" of a management practice, they are used to:

provide information on “real world” use and practice;
detect signals about the benefits and risks of...[the] use [of practices] in the general population;
help formulate hypotheses to be tested in subsequent experiments;
provide part of the community-level data needed to design more informative pragmatic clinical trials; and
inform clinical practice.

These tend to be cheaper than RCTs per unit effort.

## Nigeria data {#NG-data}

```{r}
# install.packages(c("downloader","rgdal","raster","arm","leaflet","htmlwidgets")), dependencies=TRUE)
suppressPackageStartupMessages({
  require(downloader)
  require(rgdal)
  require(sp)
  require(raster)
  require(arm)
  require(leaflet)
  require(htmlwidgets)
})
```

```{r}
# Data downloads -----------------------------------------------------------
# download IITA/OCP yield data
download("https://www.dropbox.com/s/hi75cnp3ejr4srk/OCP_trials.zip?raw=1", "OCP_trials.zip", mode = "wb")
unzip("OCP_trials.zip", overwrite = T)
sites <- read.table("sites.csv", header=T, sep=",")
trial <- read.table("trials.csv", header=T, sep=",")
tresp <- merge(sites, trial, by="sid")

# download GADM-L2 shapefile (courtesy: http://www.gadm.org)
download("https://www.dropbox.com/s/y3h6l7yu00orm78/NGA_adm2.zip?raw=1", "NGA_adm2.zip", mode = "wb")
unzip("NGA_adm2.zip", overwrite = T)
shape <- shapefile("NGA_adm2.shp")

# download raster stack (note this is a big 800+ Mb download)
download("https://www.dropbox.com/s/u5fyjbujf0d7q43/NG_250m_2017.zip?raw=1", "NG_250m_2017.zip", mode = "wb")
unzip("NG_250m_2017.zip", overwrite = T)
glist <- list.files(pattern="tif", full.names = T)
grids <- stack(glist)

# Data setup --------------------------------------------------------------
# attach GADM-L2 admin unit names from shape
coordinates(tresp) <- ~lon+lat
projection(tresp) <- projection(shape)
gadm <- tresp %over% shape
tresp <- as.data.frame(tresp)
tresp <- cbind(gadm[ ,c(5,7)], tresp)
colnames(tresp) <- c("state","lga","sid","lat","lon","alt","team","trt","ccob","tcob","twgt","cyld","tyld","ayld")

# project survey coords to grid CRS
tresp.proj <- as.data.frame(project(cbind(tresp$lon, tresp$lat), "+proj=laea +ellps=WGS84 +lon_0=20 +lat_0=5 +units=m +no_defs"))
colnames(tresp.proj) <- c("x","y")
tresp <- cbind(tresp, tresp.proj)
coordinates(tresp) <- ~x+y
projection(tresp) <- projection(tresp)

# extract gridded variables at survey locations
trespgrid <- extract(grids, tresp)
gsdat <- as.data.frame(cbind(tresp, trespgrid)) 
# plot(alt~MDEM, gsdat) ## gps altitude/location check against MDEM 
```

```{r}
# Classify by site indices ------------------------------------------------
si.lmer <- lmer(log(tyld)~trt*+(1|sid), gsdat) ## random intercept (site-level) model
display(si.lmer)
si.ran <- ranef(si.lmer) ## extract random effects
si <- as.data.frame(rownames(si.ran$sid))
si$si <- si.ran$sid[,1]
colnames(si) <- c("sid","si")
si$sic <- ifelse(si$si > 0, "A", "B") ## classify above/below average site indices (sic = A/B)
gsdat <- merge(gsdat, si, by="sid")
si <- merge(si, sites, by="sid")

# Plots
boxplot(tyld~trt, notch=T, ylab="Cob yield (kg/ha)", ylim=c(0,8000), gsdat) ## treatment differences
boxplot(tyld~sic, notch=T, ylab="Cob yield (kg/ha)", ylim=c(0,8000), gsdat) ## yield differences between site index classes
boxplot(tcob~trt*sic, notch=T, ylab="Number of cobs", ylim=c(0,800), gsdat) ## treatment differences
boxplot(tyld~trt*sic, notch=T, ylab="Cob yield (kg/ha)", ylim=c(0,8000), gsdat) ## treatment differences
plot(tyld~cyld, xlab="Maize yield (kg/ha), circular plot", ylab="Maize yield (kg/ha), total plot", gsdat)

# extract gridded variables at trial locations
si.proj <- as.data.frame(project(cbind(si$lon, si$lat), "+proj=laea +ellps=WGS84 +lon_0=20 +lat_0=5 +units=m +no_defs"))
colnames(si.proj) <- c("x","y")
si <- cbind(si, si.proj)
coordinates(si) <- ~x+y
projection(si) <- projection(tresp)
sigrid <- extract(grids, si)
sidat <- as.data.frame(cbind(si, sigrid)) 
```

# Write data frames -------------------------------------------------------
dir.create("Results", showWarnings = F)
write.csv(gsdat, "./Results/OCP_gsdat.csv", row.names = F)
write.csv(sidat, "./Results/OCP_sidat.csv", row.names = F)

# Yield trial map widget --------------------------------------------------
w <- leaflet() %>% 
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  addCircleMarkers(si$lon, si$lat, clusterOptions = markerClusterOptions())
w ## plot widget 
saveWidget(w, 'OCP_trials.html', selfcontained = T) ## save widget
```

## Central-Nigeria Small Area Estimates (Maize yield under different treatments)

