# Spatial predictions of cropland management responses {#mrt}

*Edited by: Jordan Chamberlin, Markus Walsh, ...?*

## Introduction {#mrt-intro}

An ability to track management changes and anticipate their attendant impacts requires reliable spatial information about the spatial distribution of cropland productivity and their potentials for improvement.

The most informative way of evaluating the efficacy and impact of cropland management interventions is to test these in **Multilocation Trials** (MLTs). MLTs compare interventions(s) against at least one common control, at the same geographical location and point in time. *"Multilocation"* in this context simply means having multiple individual locations (sometimes hundreds) that are subjected to the same controls and interventions that are selected randomly from within a given geographical region of interest (ROI). Data collected in multilocation trials are complex and have three fundamental aspects:

1. environmental (space-time) patterns
2. nonstructural (space-time) noise
3. relationships among genotypes, management actions and environments.

The main problems with all MLTs in agriculture is that they are usually logistically difficult and quite expensive to conduct *at scale*. MLTs at scale typically require collaborative working approaches and models.

The second scale dependent issue relates to how locations are defined.

This chapter provides ...

## Nigeria fertilizer response data {#NG-data}

```{r}
# install.packages(c("downloader","rgdal","raster","arm","leaflet","htmlwidgets")), dependencies=TRUE)
suppressPackageStartupMessages({
  require(downloader)
  require(rgdal)
  require(sp)
  require(raster)
  require(arm)
  require(leaflet)
  require(htmlwidgets)
})
```

```{r}
# Data downloads -----------------------------------------------------------
# download IITA/OCP yield data
download("https://www.dropbox.com/s/hi75cnp3ejr4srk/OCP_trials.zip?raw=1", "OCP_trials.zip", mode = "wb")
unzip("OCP_trials.zip", overwrite = T)
sites <- read.table("sites.csv", header=T, sep=",")
trial <- read.table("trials.csv", header=T, sep=",")
tresp <- merge(sites, trial, by="sid")

# download GADM-L2 shapefile (courtesy: http://www.gadm.org)
download("https://www.dropbox.com/s/y3h6l7yu00orm78/NGA_adm2.zip?raw=1", "NGA_adm2.zip", mode = "wb")
unzip("NGA_adm2.zip", overwrite = T)
shape <- shapefile("NGA_adm2.shp")

# download raster stack (note this is a big 800+ Mb download)
download("https://www.dropbox.com/s/u5fyjbujf0d7q43/NG_250m_2017.zip?raw=1", "NG_250m_2017.zip", mode = "wb")
unzip("NG_250m_2017.zip", overwrite = T)
glist <- list.files(pattern="tif", full.names = T)
grids <- stack(glist)

# Data setup --------------------------------------------------------------
# attach GADM-L2 admin unit names from shape
coordinates(tresp) <- ~lon+lat
projection(tresp) <- projection(shape)
gadm <- tresp %over% shape
tresp <- as.data.frame(tresp)
tresp <- cbind(gadm[ ,c(5,7)], tresp)
colnames(tresp) <- c("state","lga","sid","lat","lon","alt","team","trt","ccob","tcob","twgt","cyld","tyld","ayld")

# project survey coords to grid CRS
tresp.proj <- as.data.frame(project(cbind(tresp$lon, tresp$lat), "+proj=laea +ellps=WGS84 +lon_0=20 +lat_0=5 +units=m +no_defs"))
colnames(tresp.proj) <- c("x","y")
tresp <- cbind(tresp, tresp.proj)
coordinates(tresp) <- ~x+y
projection(tresp) <- projection(tresp)

# extract gridded variables at survey locations
trespgrid <- extract(grids, tresp)
gsdat <- as.data.frame(cbind(tresp, trespgrid)) 
# plot(alt~MDEM, gsdat) ## gps altitude/location check against MDEM 
```

```{r}
# Classify by site indices ------------------------------------------------
si.lmer <- lmer(log(tyld)~trt*+(1|sid), gsdat) ## random intercept (site-level) model
display(si.lmer)
si.ran <- ranef(si.lmer) ## extract random effects
si <- as.data.frame(rownames(si.ran$sid))
si$si <- si.ran$sid[,1]
colnames(si) <- c("sid","si")
si$sic <- ifelse(si$si > 0, "A", "B") ## classify above/below average site indices (sic = A or B)
gsdat <- merge(gsdat, si, by="sid")
si <- merge(si, sites, by="sid")

# Plots
boxplot(tyld~trt, notch=T, ylab="Cob yield (kg/ha)", ylim=c(0,8000), gsdat) ## treatment differences
boxplot(tyld~sic, notch=T, ylab="Cob yield (kg/ha)", ylim=c(0,8000), gsdat) ## yield differences between site index classes
boxplot(tcob~trt*sic, notch=T, ylab="Number of cobs", ylim=c(0,800), gsdat) ## treatment differences
boxplot(tyld~trt*sic, notch=T, ylab="Cob yield (kg/ha)", ylim=c(0,8000), gsdat) ## treatment differences
plot(tyld~cyld, xlab="Maize yield (kg/ha), circular plot", ylab="Maize yield (kg/ha), total plot", gsdat)

# extract gridded variables at trial locations
si.proj <- as.data.frame(project(cbind(si$lon, si$lat), "+proj=laea +ellps=WGS84 +lon_0=20 +lat_0=5 +units=m +no_defs"))
colnames(si.proj) <- c("x","y")
si <- cbind(si, si.proj)
coordinates(si) <- ~x+y
projection(si) <- projection(tresp)
sigrid <- extract(grids, si)
sidat <- as.data.frame(cbind(si, sigrid)) 
```

```{r}
# Write data frames -------------------------------------------------------
dir.create("Results", showWarnings = F)
write.csv(gsdat, "./Results/OCP_gsdat.csv", row.names = F)
write.csv(sidat, "./Results/OCP_sidat.csv", row.names = F)
```

```{r}
# Yield trial map widget --------------------------------------------------
w <- leaflet() %>% 
  addProviderTiles(providers$OpenStreetMap.Mapnik) %>%
  addCircleMarkers(si$lon, si$lat, clusterOptions = markerClusterOptions())
w ## plot widget 
saveWidget(w, 'OCP_trials.html', selfcontained = T) ## save widget
```

## Western Kenya One Acre Fund monitoring survey {WK-data}
